Case based Arabic Morphological Analysis with Compositional Non-deterministic Automata

### 2010-06-25 tex/sarf.tex
added sarf section

@@

\section{Background}

% nondeterministic finite-state automata

\section{Related Work }

@@

Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with 
proper names and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
as well as biblical sources~\cite{}.

\subsection{Sarf example }
\label{sec:example}

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/FSM.pstex_t}}
\caption{Example composition of the prefix, stem and suffix FSAs}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words

@@

\subsection{Recursive Affixes}
% smaller FSM
% retain POS info

\subsection{Affix Linear FSM}

\subsection{Stem Linear FSM}

@@

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/abstract_machine.pstex_t}}
\caption{An abstraction of the non-deterministic composition of the prefix, stem and suffix FSAs}
\label{f:example}
}
\end{figure*}

% algorithm
% diagram of abstract machines

\begin{Verbatim}[fontsize=\relsize{-1},
frame=topline,framesep=4mm,label=\fbox{NDFSA Sarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
Sarf(string $L$, Controller $A$) 
  FSAVec $M$; -- collection of FSA automata
  FSA $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      $A$.control($M$, $c$);
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhiteORNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} \}
\end{Verbatim}

\section{Islamic Literature Case Study}

\subsection{Hadith Segmentation}

\subsection{Chain of Narrators}

\subsection{Controller}
% FSM for the controller

>>>>

\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}

@@

%\section{Background}
% TODO: nondeterministic finite-state automata

\section{Related Work }
\label{sec:related}

@@

Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with 
proper names and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.

Sarf encodes the prefix, suffix and stem lexicons into 
three linear FSAs and uses a non-deterministic composition
of the three FSAs to compute all valid morphological analyses.
We first present a runing example of Sarf. 
The running example explains
how the non-deterministic composition works and 
illustrates how Sarf implements recursive affixes and
handles ``run-on'' words.
Then we discuss how Sarf builds and composes the different FSAs.

\subsection{Sarf example}
\label{sec:example}

\begin{figure*}[tb]
\center{
\resizebox{1.9\columnwidth}{!}
{ \input{psfigs/FSM.pstex_t}}
\caption{Example composition of the prefix, stem and suffix FSAs.}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words

@@

\subsection{Recursive Affixes}
\label{sec:recaffix}

Morphological analysis partitions a given Arabic word
into a set of morphemes.
A morpheme is either an affix or a stem. 
We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\cite{Buckwalter:02} keeps separate lists 
of prefixes, suffixes and stems and assigns a category
to each morphemes. 
The relation between morphemes and categories is one 
to many.

A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The string $\alpha\beta\gamma$ partition of the string is a 
valid morphological analysis if  and only if we have
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
Moreover, the POS tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping these morphemes separate in a list
is redundant and may produce consistency issues. 
We add two lists of compatibilty rules $L_{pp}$ and
$L_{xx}$ for prefix-prefix and suffix suffix compatibility
respectively.
This allows us to keep a shorter list of affixes. 
We need one more additional modification to accomodate 
a category for the morphemes accepted by the $L_{pp}$
and the $L_{xx}$ rules. For that we introduce the concept
of a {\em resulting category}. 
The resulting category in all practical cases
was the category of the second morpheme in the pair.

\subsection{Affix Linear FSA}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\cite{Buckwalter:02}
and extracted our refined lists of recursives affixes.
We encoded the lists into a FSA similar to that
in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
The two FSAs are directed acyclic graphs. 
The abscence of cycles in these FSAs reduces the 
traversal of the affix FSAs to a linear traversal.
This is computationaly superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The prefix and suffix FSAs encode the $L_{pp}$ and
$L_{xx}$ compatibility lists.
An accept state in the FSA corresponds to the last character
of a prefix in the predix and suffix lists.
The accept states hold category state information as well
as POS and other tags.

\subsection{Stem Linear FSA}
\label{sec:stemFSA}

We built our stem lexicon using the stem lexicon of 
Buckwalter. 
We augmented the lexicon with a set of proper names and
a set of location names. 
We obtained the set of proper names from online 
and biblical sources.

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is also a linear FSA where the accept
states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since again it is
a linear walk in the trie while with Buckwalter it is 
several hash lookups.

@@

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic composition 
of the prefix, stem and suffix FSA. 
The circle represents regular states, the square
represents accept state and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix FSAs
are all linear and do not have cycles as discussed 
is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
The cycles are introduced in the abstract machine
for brevity and illustrative purposes.

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/abstract_machine.pstex_t}}
\caption{An abstraction of the non-deterministic composition of 
    the prefix, stem and suffix FSAs.}
\label{f:composition}
}
\end{figure*}

The algorithm \CodeIn{NDSarf} below controls Sarf. 
It takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one FSA $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
of FSAs $M$. 
The algorithm reads a character $c$ at a time from $L$
and moves on all the machines in $M$. 
For each $\Phi$ in $M$ the algorithm checks whether
$\Phi$ is in an accept state or not.

If $\Phi$ is in an accpet state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last character leading to the current state
was a non-connecting character.

In all cases, if $\Phi$ is in an accept state, 
it spawns another FSA $\Psi$ and adds it to $M$. 
If $c$ leads to a reject state, then $\Phi$ dies 
and we remove it from $M$. 
Otherwise, $\Phi$ transitions using the $c$ edge.
We invoke the controller $A$ after all the FSAs in $M$ 
have moved and allow it full access and control over
all FSAs in $M$.

The arrows in Figure~\ref{f:composition} that transition
from prefix to stem, stem to suffix, and suffix to prefix
show the compatibility constraints as a condition for 
the transition.

\begin{table}[tb]
\centering
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-1},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  FSAVec $M$; -- collection of FSA automata
  FSA $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} 
    $A$.control($M$, $c$); \}
\end{Verbatim}
\end{tabular}
\label{a:ndsarf}
\end{table}

\section{Islamic Literature Case Study}
\label{sec:islamic}

A hadith in Islamic literature is a narration from the prophet Mohamad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to sucessfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. 
  Our understanding is that those books are historical documents with 
  public IP.}.

Our analysis accpets a book as a text stream input
and segements it into a vector of individual hadiths. 
We segment each hadith into two parts. The first part
is the sanad and it
contains the chain of narrators who related the hadith
to Mohamad. 
The second part is the matn and it constitutes the content
of the hadith. 
We are concerned in further exploring the sanad. 
In particular, we detect the chain of narrators
and the relation between each narrator, his ancestor, and 
his predecessor in the chain. 
We also detect the full name of each narrator that is
usually composed of several proper names with connectors
in between.

We built our case-based controller to target
the detection of proper names. 
This includes the task of finding compound names 
composed of several words such as \RL{`bd alr.hmn} often
appearing as ``run-on'' words.
The controller also targets words that mean narrate when
they appear in the neighborhood of multiple proper names.

\subsection{Controller}
% FSM for the controller
The FSA in Figure~\ref{f:hadith}
